---
title: "How to use DiagrammeR+Mermaid?"
subtitle: "Short answer: Don't!"
author: "José R. Ferrer-Paris"
date: 01 Jul 2024
categories:
 - Mermaid.js
 - Diagrams
execute: 
  echo: true
---

## What I want to do

I want to use the R package `DiagrammeR` to customise a Mermaid.js diagram.

## Challenges

I learned to use Mermaid.js to embed reproducible diagrams in quarto documents. Mermaid.js in under active development, syntax and features change from version to version. 

Usually Mermaid code can be embeded in a quarto document like this:

```{mermaid}
%%| echo: fenced
%%| label: fig-mermaid
%%| fig-cap: A mermaid graph
graph LR
  A[qmd] --> B(Knitr)
  A[qmd] --> C(Jupyter)
  B(Knitr) --> D[md]
  C(Jupyter) --> D[md]
  D[md] --> E(pandoc)
  E(pandoc) --> F(HTML)
  E(pandoc) --> G(PDF)
  E(pandoc) --> H(Word)
  E(pandoc) --> I{and more}
```

:::{.aside}
This is the same diagramm code as the figure shown [here](/mark-updown.qmd).
:::

However, when loading the library `DiagrammeR` in a document, it will call a fixed version of mermaid. This means that it is stuck with a slightly outdated syntax, and that usually means sacrificing some features. Mixing `mermaid` code blocks and diagrams created within R code blocks in the same document will have unexpected results.

I am slightly inclined _against_ using `DiagrammeR` for mermaid plots, but I am still testing options.

## Sources

- https://rdrr.io/cran/DiagrammeR/src/R/mermaid.R
- https://www.reddit.com/r/github/comments/100mbuh/mermaid_diagram_not_working/
- https://stackoverflow.com/questions/58689080/mermaid-diagrams-not-rendering-correctly-in-rmarkdown-xaringan-presentations

## Steps in R

Load the libraries:

```{r libimport}
library(DiagrammeR)
library(widgetframe)
```

```{r}
example1 <- DiagrammeR::mermaid("
graph LR;
A((Orange)) --> B((Grey));

classDef orange fill:#f96;
classDef grey fill:#d3d3d3;
class A orange;
class B grey;
")

```

```{r}
#| fig-cap: |
#|   How Quarto orchestrates rendering of documents: start with 
#|   a qmd file, use the Knitr or Jupyter engine to perform the 
#|   computations and convert it to an md file, then use Pandoc 
#|   to convert to various file formats including HTML, PDF, 
#|   and Word.
example2 <- DiagrammeR("
graph LR
  A[qmd] --> B(Knitr)
  A[qmd] --> C(Jupyter)
  B(Knitr) --> D[md]
  C(Jupyter) --> D[md]
  D[md] --> E(pandoc)
  E(pandoc) --> F(HTML)
  E(pandoc) --> G(PDF)
  E(pandoc) --> H(Word)
  E(pandoc) --> I{and more}"
     )
```


```{r attempt2}
example3 <- mermaid("
sequenceDiagram
  participant ParticipantA
  participant ParticipantB
  participant ParticipantC
  ParticipantA->>ParticipantB: I want something
  ParticipantB->>ParticipantC: he want something
  ParticipantC->>ParticipantB: here is something
  ParticipantB->>ParticipantA: he got something for you
  ", height = '100%', width = '100%')
```


Direct output (no `<iframe>`):

```{r}
#| fig-height: 2
example1
# same as: widgetframe::frameableWidget(example1)
```

With an `<iframe>` created with `widgetframe`:
```{r}
#| fig-height: 2
widgetframe::frameWidget(example1, height=200)
```


```{r}
widgetframe::frameWidget(example2, height=300)
```

```{r}
widgetframe::frameWidget(example3, height=500)
```